---
title: "Grain-size tables"
subtitle: "The sediment fraction, the <63 µm fines, and the fraction distribution"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Step 3 records the sediment fraction each measurement sits on (bulk or sieved) and
moves the grain-size distribution into its own **`grain_size_fraction`** table. The
per-subsample `<63 µm` fines (`fines_lt63`) and a fraction summary
(`frac_class` / `sieve_um`) live on **`subsample`**; the authoritative per-value
fraction is on `measurement`. 4Demon has no grain-size distribution, but does carry
the sediment fraction.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)

db_dir <- "data/db"
all_srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  stem   = c("mareano", "vannmiljo", "ices_dome", "mudab", "4demon"))
gsrcs <- all_srcs[all_srcs$stem != "4demon", ]   # sources with grain-size

qry <- function(stem, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_clean.sqlite")))
  on.exit(dbDisconnect(con)); as_tibble(dbGetQuery(con, sql))
}
kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
```

## 1. Sediment fraction (bulk / sieved / mixed)

Whether a subsample's target metals were measured on bulk (whole sediment or
`<2 mm`) or a sieved fine fraction. `mixed` marks the subsamples whose targets span
both fractions; `None` means the subsample has no target chemistry to classify. The
per-measurement `frac_class` + `sieve_um` on `measurement` give the exact fraction
of every value.

```{r}
bind_rows(lapply(seq_len(nrow(all_srcs)), function(i)
  qry(all_srcs$stem[i],
      "SELECT COALESCE(target_frac_class,'None') target_frac_class, COUNT(*) n
       FROM subsample GROUP BY target_frac_class") |>
    mutate(Source = all_srcs$Source[i]))) |>
  pivot_wider(names_from = target_frac_class, values_from = n, values_fill = 0) |>
  select(Source, any_of(c("bulk", "sieved", "mixed", "None"))) |>
  kfmt()
```

## 2. The <63 µm fines

`fines_lt63` is the percentage of material finer than 63 µm (the clay + silt
"mud"), carried on `subsample` for the sources with grain-size. Mud-dominated
samples are those at least 50 % fines.

```{r}
bind_rows(lapply(seq_len(nrow(gsrcs)), function(i)
  qry(gsrcs$stem[i], "SELECT
        SUM(fines_lt63 IS NOT NULL) with_fines,
        SUM(fines_lt63 >= 50) mud_dominated,
        ROUND(AVG(fines_lt63),1) mean_fines
      FROM subsample") |>
    mutate(Source = gsrcs$Source[i]))) |>
  transmute(Source, `Subsamples with fines` = with_fines,
            `Mud-dominated (>=50%)` = mud_dominated, `Mean fines (%)` = mean_fines) |>
  kfmt()
```

## 3. Fraction distribution

The `grain_size_fraction` table holds the individual size fractions (corrected
percentages, with parsed `lo_um`/`hi_um` bounds) moved out of `measurement`.

```{r}
bind_rows(lapply(seq_len(nrow(gsrcs)), function(i)
  qry(gsrcs$stem[i], "SELECT COUNT(*) rows, COUNT(DISTINCT subsample_id) subsamples,
                      COUNT(DISTINCT symbol) codes FROM grain_size_fraction") |>
    mutate(Source = gsrcs$Source[i]))) |>
  transmute(Source, `Fraction rows` = rows, `Subsamples` = subsamples,
            `Distinct codes` = codes) |>
  kfmt()
```

::: {.callout-note}
`fines_lt63` and the fractions are derived, not deleted-from, quantities; the
grain-size correction (renormalising scaled curves, flagging unreliable values)
was applied in the slim stage, documented under
[Grain size](grain-size-fraction.qmd).
:::
