---
title: "Clean schema"
subtitle: "The tables and columns of the clean databases"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Each `\<source\>_clean.sqlite` has the seven shared tables carried from the slim
schema (`element`, `dataset`, `site`, `event`, `subsample`, `measurement`,
`method`), plus, where the source has the data, `organic_carbon` and
`grain_size_fraction`. 4Demon has neither (no organic carbon, no grain-size). The
main differences from the slim schema: `measurement` holds only the target
elements and the Fe/Al normalisers, with aggregated values and a bulk/sieved
fraction label; organic carbon and grain-size sit in their own tables; and depths
are in cm.

```{r}
library(tibble)
library(knitr)
```

## measurement (target elements + Fe/Al)

One row per cleaned measurement of a target element or a Fe/Al normaliser. Failed-QC
rows are gone; duplicates and technical replicates are collapsed. Organic carbon and
grain-size composition have moved to their own tables.

```{r}
tribble(
  ~Column,        ~Type,    ~Meaning,
  "measurement_id","INTEGER","Primary key (carried from slim).",
  "subsample_id",  "INTEGER","-> subsample.",
  "symbol",        "TEXT",   "Element, ICES name (-> element).",
  "value",         "REAL",   "Value in the original (ICES-named) unit.",
  "unit",          "TEXT",   "Original reporting unit.",
  "value_std",     "REAL",   "Standardised value, mg/kg (mean where replicates were averaged).",
  "unit_std",      "TEXT",   "Always mg/kg.",
  "value_sd",      "REAL",   "SD across technical replicates, else NULL.",
  "n_rep",         "INTEGER","Number of distinct replicate values averaged (1 = single).",
  "value_uncrt",   "REAL",   "Analytical uncertainty, mg/kg (ICES-DOME only), else NULL.",
  "frac_class",    "TEXT",   "Sediment fraction measured: 'bulk' (whole sample or <2mm) or 'sieved' (a fine sub-fraction).",
  "sieve_um",      "REAL",   "Sieve cutoff in um for sieved rows (e.g. 63, 20), NULL for bulk.",
  "method_id",     "INTEGER","-> method."
) |> kable()
```

`frac_class` / `sieve_um` are derived from the source's sediment-fraction code (the
ICES `matrix`), which is not carried forward. Bulk versus sieved is recorded per
measurement because it varies within a subsample: a sample's metals and its organic
carbon, and even different target metals, are often measured on different fractions.

## organic_carbon

One row per organic carbon measurement (CORG, TOC, TOC63), split out of
`measurement` as supplementary (not a main analyte). Same columns as `measurement`.
Absent for 4Demon.

## subsample

One row per subsample (a depth slice of a sampling event), with depth, the
supporting-data flags, a per-subsample sediment-fraction summary, and the mud
content.

```{r}
tribble(
  ~Column,       ~Type,     ~Meaning,
  "subsample_id","INTEGER", "Primary key.",
  "event_id",    "INTEGER", "-> event.",
  "depth_from",  "REAL",    "Top of the slice, cm.",
  "depth_to",    "REAL",    "Bottom of the slice, cm.",
  "fe_exist",    "INTEGER", "1/0: an Fe value exists for this subsample.",
  "al_exist",    "INTEGER", "1/0: an Al value exists.",
  "org_exist",   "INTEGER", "1/0: an organic carbon value exists.",
  "comp_exist",  "INTEGER", "1/0: grain-size composition exists.",
  "frac_class",  "TEXT",    "Fraction summary of the subsample's target measurements: 'bulk', 'sieved', or 'mixed' (targets span both). NULL if no target chemistry.",
  "sieve_um",    "REAL",    "Target sieve cutoff where unambiguous, else NULL.",
  "fines_lt63",  "REAL",    "% finer than 63 um (clay + silt mud content), else NULL.",
  "fines_basis", "TEXT",    "How fines_lt63 was derived."
) |> kable()
```

The authoritative bulk/sieved value is per measurement; `subsample.frac_class` is a
convenience summary, with `mixed` marking the (few) subsamples whose target metals
span both fractions.

## grain_size_fraction (per fraction measurement)

The grain-size distribution moved out of `measurement`. One row per grain-size
mass-fraction value. Absent for 4Demon.

```{r}
tribble(
  ~Column,       ~Type,    ~Meaning,
  "subsample_id","INTEGER","-> subsample.",
  "symbol",      "TEXT",   "The source's grain-size code.",
  "matrix",      "TEXT",   "Sample fraction the value is a percentage of (where recorded).",
  "lo_um",       "REAL",   "Lower size bound, um.",
  "hi_um",       "REAL",   "Upper size bound, um (NULL = open, e.g. gravel).",
  "value_pct",   "REAL",   "Corrected mass fraction, %."
) |> kable()
```

## Other tables

`element` (symbol, `name`, category, `cas`), `dataset`, `site` (coordinates plus a
`depth` column), `event` (date, year, descriptive `sampling_tool`; `datetime` /
`time` dropped), and `method` (technique, `method_description`, lab, `lod` / `loq`
with `limit_unit`) carry over from slim with the harmonisations applied.

## What changed from slim

```{r}
tribble(
  ~Change,                              ~Detail,
  "Rows removed",                       "Out of range, invalid, below limit, wet weight, source-specific QC.",
  "Duplicates / replicates collapsed",  "One value each; replicates gain value_sd + n_rep.",
  "measurement is targets + Fe/Al",     "Organic carbon to organic_carbon; grain-size composition to grain_size_fraction.",
  "Sediment fraction as frac_class",    "The matrix code becomes frac_class (bulk/sieved) + sieve_um; a bulk/sieved/mixed summary is added to subsample.",
  "Flag columns dropped",               "The slim QC flags were consumed by the cleaning and are not carried.",
  "Depths in cm",                       "ICES-DOME converted from metres; Vannmiljo corrupt depths nulled.",
  "Symbols / units unified",            "ICES names; TOC -> CORG; value_std in mg/kg; CAS added to element."
) |> kable()
```
